---
title: "JONATHAN_2027_POLLING_INTELLIGENCE_REPORT"
author: "Owen Bolongo"
date: "2026-05-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir="C:/Users/user/Desktop/May-August_Workshop-Practice/JONATHAN_2027_POLLING_INTELLIGENCE_REPORT")
```


##Step 3: Install the necessary packages
```{r}
#install.packages("readxl")#Reading excel files
#install.packages("tidyverse")#dplyr, ggplot, tidyr, stringr
#install.packages("janitor")#Cleaning Messy Colums
#install.packages("skimr")# Quick data summary
#install.packages("scales")# Formatting axes in ggplot2
#install.packages("ggthemes")#Provide extra ggplot themes
#install.packages("viridis")#color-blind-friendly color scales
#install.packages("corrplot")#correlation matrix plot
#install.packages("caret")#Machine Learning Frameworks
#install.packages("randomForest")# Random Forest model
#install.packages("cluster") #K-means Cluster
#install.packages("factoextra")#Visualizing Clustering results
#install.packages("leaflet") #interactive maps
```

#Step 4: Loading the necessary library
```{r}
library(readxl)
library(tidyverse)
library(janitor)
library(skimr)
library(ggthemes)
library(viridis)
library(corrplot)
library(randomForest)
#library(factoextra)
library(leaflet)
```

#Step 5: Load the dataset
```{r}
df_raw<-read_excel("KITUI.xlsx")
```

##Step 6: View the dataset/Look at the dataset
```{r}
#View the first 6 observations
      #head(df_raw)
      #View(df_raw)
      #print(df_raw)

#View the last 6 rows of the observations
     #tail(df_raw)

#Get dimmensions: number of rows * columns
     #dim(df_raw)

#Get the column names
     #colnames(df_raw)
     #names(df_raw)
#Get the datatypes and structure
     #str(df_raw)

```

#Step 7: Data cleaning and Visualization
```{r}
df<-df_raw%>%
  clean_names()
  colnames(df)
#View(df)
```

```{r}
df<-df%>%
  mutate(charity_kaluki_ngilu=ifelse(
        constituency_code==69 & 
        constituency_name=="MWINGI CENTRAL" & 
        polling_station_code=="15069034001401" &             
        polling_station_name=="YANYONGE NURSERY SCHOOL",
        0,
        charity_kaluki_ngilu
  ))

df<-df%>%
  mutate(charity_kaluki_ngilu=ifelse(
        constituency_code==69 & 
        constituency_name=="MWINGI CENTRAL" & 
        polling_station_code=="15069034101801" &             
        polling_station_name=="KANYUNGA DISPENSARY",
        0,
        charity_kaluki_ngilu
  ))

df<-df%>%
  mutate(charity_kaluki_ngilu=ifelse(
        constituency_code==69 & 
        constituency_name=="MWINGI CENTRAL" & 
        polling_station_code=="15069034205101" &             
        polling_station_name=="KYANGU PRIMARY SCHOOL",
        0,
        charity_kaluki_ngilu
  ))



```
#Step 8: Checking for missing values
```{r}
#Count NAs Per Column
     #colSums(is.na(df))
#More detailed:Which rows have any NA ?
df<-df%>%
  filter(!(if_any(everything(),is.na)))
nrow(df)
```

#Step 9: Finding distinct value
```{r}
df<-df%>%
  distinct(constituency_code,constituency_name,polling_station_code,polling_station_name, .keep_all = TRUE)
#View(df)
```

#Step: 10 Standardize constituency names
```{r}
#Check unique constituency names
unique(df$constituency_name)

#TRIM WHITE SPACES AND CONVERT THEM TO UPPERCASE
df<-df%>%
  mutate(
    constituency_name=str_trim(constituency_name),
    constituency_name=str_to_upper(constituency_name)
  )
```

#Step 11: Validate Vote Arithmetic
A critical Validation:Valid votes must be equal to votes casted minus rejected votes. Any discrepancy signals data enty error
```{r}
df<-df%>%
  mutate(
    calculated_valid=votes_cast-rejected_votes,
    vote_check = calculated_valid == valid_votes 
    )
```


#Step:12 Get the vote share of each individual as percentage of the total votes
```{r}
df<-df%>%
  mutate(
    mueke_share=(jonathan_mwangangi_mueke/valid_votes)*100,
    malombe_share=(julius_makau_malombe/valid_votes)*100,
    musila_share=(david_musila/valid_votes)*100,
    ngilu_share=(charity_kaluki_ngilu/valid_votes)*100,
    # Turnout rate (% of registered voters who voted — approximate) 
    rejection_rate = (rejected_votes / votes_cast) * 100,
    
     # Mueke's margin vs winner (positive = Mueke is winning) 
    mueke_vs_malombe = jonathan_mwangangi_mueke - julius_makau_malombe, 
    mueke_vs_musila  = jonathan_mwangangi_mueke - david_musila,
    
     # Who won this polling station? 
    winner = case_when( 
        julius_makau_malombe== pmax(julius_makau_malombe,jonathan_mwangangi_mueke, 
                                      david_musila, charity_kaluki_ngilu) ~ 'Malombe', 
      jonathan_mwangangi_mueke == pmax(julius_makau_malombe,jonathan_mwangangi_mueke, 
                                      david_musila, charity_kaluki_ngilu) ~ 'Mueke', 
      david_musila== pmax(julius_makau_malombe,jonathan_mwangangi_mueke, 
                                      david_musila, charity_kaluki_ngilu) ~ 'Musila', 
      TRUE ~ 'Ngilu' 
    ) 
    
  )
#View(df)
```

#Step 2.7 — Final Clean Dataset
```{r}
# Remove the validation helper columns, keep the clean dataset 
df_clean <- df %>% 
  select(-calculated_valid, -vote_check) 
 
# Save as CSV for reproducibility 
write.csv(df_clean, 'kitui_clean.csv', row.names = FALSE) 
 
# Quick final overview 
skim(df_clean) 
 
# Explanation: 
# select(-col)   removes a column (the minus sign means exclude) 
# write.csv()    exports to CSV format for sharing 
# skim()         prints a rich summary: mean, SD, missing, histograms 
```

#WEEK 3 — County-Level Performance Analysis
##Step 3.1 — Total Votes by Candidate 
```{r}
# Aggregate total votes for each candidate across all stations 
county_totals <- df_clean %>% 
  summarise( 
    Malombe = sum(julius_makau_malombe), 
    Mueke   = sum(jonathan_mwangangi_mueke), 
    Musila  = sum(david_musila), 
    Ngilu   = sum(charity_kaluki_ngilu), 
    Total_Valid = sum(valid_votes) 
  ) %>% 
  pivot_longer( 
    cols = c(Malombe, Mueke, Musila, Ngilu), 
    names_to  = 'candidate', 
    values_to = 'votes' 
  ) %>% 
  mutate(vote_share = (votes / sum(votes)) * 100) 

View(county_totals)
 
# Explanation: 
# summarise()      collapses all rows into summary statistics 
# sum()            adds up all values in a column 
# pivot_longer()   converts wide format to long format 
#                  (one row per candidate, needed for ggplot2) 
# cols = c(...)    which columns to pivot 
# names_to         new column for old column names 
# values_to        new column for old column values 
```

#Step 3.2 — Bar Chart: Total Votes by Candidate 
The foundation of ggplot2 is the grammar of graphics: data + aesthetics + geometry. Every ggplot2 chart follows 
this structure:

```{r}
ggplot(data = county_totals,             # 1. DATA: which data frame to use 
       aes(x = reorder(candidate, -votes),# 2. AESTHETICS: x axis = candidate 
           y = votes,                     #               y axis = vote count 
           fill = candidate)) +           #               fill color = candidate 
  geom_col(width = 0.6) +                 # 3. GEOMETRY: draw bars 
  geom_text(aes(label = scales::comma(votes)),  # 4. Add number labels 
            vjust = -0.5, size = 4, fontface = 'bold') + 
  scale_fill_manual(values = c(           # 5. Custom colors 
    'Malombe' = '#1B3A6B', 
    'Mueke'   = '#0EA5E9', 
    'Musila'  = '#16A34A', 
    'Ngilu'   = '#D97706' 
  )) + 
  scale_y_continuous(labels = scales::comma) + # Format y-axis numbers 
  labs(                                   # 6. LABELS 
    title    = '2022 Kitui Governor: Total Votes by Candidate', 
    subtitle = 'Polling station results aggregated to county level', 
    x        = 'Candidate', 
    y        = 'Total Votes', 
    caption  = 'Source: IEBC Form 37C | Kitui County' 
  ) + 
  theme_minimal(base_size = 13) +          # 7. Clean theme 
  theme(legend.position = 'none',          # Remove redundant legend 
        plot.title = element_text(face = 'bold', color = '#1B3A6B')) 
 
ggsave('week3_county_totals.png', width = 9, height = 6, dpi = 300)
```

#Step 3.3 — Pie Chart of Vote Share
```{r}
ggplot(county_totals, aes(x = '', y = vote_share, fill = candidate)) + 
  geom_col(width = 1, color = 'white', linewidth = 0.5) +  # Stacked bars 
  coord_polar(theta = 'y') +                # Convert bars to pie/donut 
  geom_text(aes(label = paste0(candidate, '\n', round(vote_share, 1), '%')), 
            position = position_stack(vjust = 0.5), 
            size = 4, color = 'white', fontface = 'bold') + 
  scale_fill_manual(values = c( 
    'Malombe' = '#1B3A6B', 'Mueke' = '#0EA5E9', 
    'Musila' = '#16A34A', 'Ngilu' = '#D97706' 
  )) + 
  labs(title = '2022 Kitui Governor: Vote Share', caption  = 'Source: IEBC Form 37C | Kitui County',
       subtitle = 'Julius Makau Malombe received 60.3% of the county vote') + 
  theme_void() +                  # Remove all axes and gridlines 
  theme(legend.position = 'right', 
        plot.title = element_text(face = 'bold', size = 14)) 
 
# Explanation: 
# coord_polar()  transforms the cartesian coordinate bar chart into a pie 
# theta = 'y'    the y axis becomes the angle 
# position_stack(vjust=0.5) places labels at the center of each segment 
# theme_void()   removes background, axes, and gridlines — ideal for pies 
```
#Step 3.4 — Constituency-Level Summary Table
```{r}
constituency_summary <- df_clean %>% 
  group_by(constituency_name) %>% 
  summarise( 
    total_stations  = n(), 
    total_votes_cast = sum(votes_cast), 
    malombe_votes   = sum(julius_makau_malombe), 
    mueke_votes     = sum(jonathan_mwangangi_mueke), 
    musila_votes    = sum(david_musila), 
    ngilu_votes     = sum(charity_kaluki_ngilu), 
    mueke_pct       = round((mueke_votes / sum(valid_votes)) * 100, 2) 
  ) %>% 
  arrange(desc(mueke_pct)) 
 
# Print to console 
View(constituency_summary) 
 
# Explanation: 
# group_by()   creates groups — all summarise() calculations happen 
#              separately for each constituency 
# n()          counts the number of rows in each group 
# arrange()    sorts the results 
# desc()       sorts in descending order (highest first)
```

#WEEK 4 — Constituency Intelligence & Heatmap Analysis 
##Step 4.1 — Faceted Bar Charts by Constituency
Faceted charts show the same visualization repeated for each group — one panel per constituency: 
```{r}
# Reshape data to long format for faceted chart 
constituency_long <- df_clean %>% 
  group_by(constituency_name) %>% 
  summarise( 
    Malombe = sum(julius_makau_malombe), 
    Mueke   = sum(jonathan_mwangangi_mueke), 
    Musila  = sum(david_musila), 
    Ngilu   = sum(charity_kaluki_ngilu) 
  ) %>% 
  pivot_longer( 
    cols = c(Malombe, Mueke, Musila, Ngilu), 
    names_to = 'candidate', 
    values_to = 'votes' 
  ) 
 
# Build faceted chart 
ggplot(constituency_long, 
       aes(x = reorder(candidate, -votes), y = votes, fill = candidate)) + 
  geom_col(width = 0.65, show.legend = FALSE) + 
  facet_wrap(~ constituency_name, scales = 'free_y', ncol = 3) + 
  scale_fill_manual(values = c( 
    'Malombe' = '#1B3A6B', 'Mueke' = '#0EA5E9', 
    'Musila'  = '#16A34A', 'Ngilu' = '#D97706' 
  )) + 
  scale_y_continuous(labels = scales::comma) + 
  labs( 
    title    = 'Vote Distribution by Constituency — 2022 Kitui Governor', 
    subtitle = 'Each panel shows vote totals for the four main candidates', 
    x = NULL, y = 'Total Votes' 
  ) + 
  theme_minimal(base_size = 11) + 
  theme( 
    strip.text       = element_text(face = 'bold', color = '#1B3A6B'), 
    strip.background = element_rect(fill = '#EFF6FF', color = NA), 
    axis.text.x      = element_text(angle = 30, hjust = 1) 
  ) 
 
# Explanation: 
# facet_wrap(~ constituency_name)  creates one panel per constituency 
# scales = 'free_y'   each panel has its own y-axis scale 
# ncol = 3            arrange panels in 3 columns 
# strip.text          styles the panel header text 
# strip.background    styles the panel header background 
# axis.text.x angle   rotates x-axis labels to prevent overlap 
```







#Step 4.2 — Heatmap of Mueke Vote Share by Constituency 
A heatmap uses color intensity to show where values are high or low — ideal for geographic political 
intelligence: 
```{r}
# Calculate vote share per candidate per constituency 
share_matrix <- df_clean %>% 
  group_by(constituency_name) %>% 
  summarise( 
    Malombe = round(sum(julius_makau_malombe) / sum(valid_votes) * 100, 1), 
    Mueke   = round(sum(jonathan_mwangangi_mueke) / sum(valid_votes) * 100, 1), 
    Musila  = round(sum(david_musila) / sum(valid_votes) * 100, 1), 
    Ngilu   = round(sum(charity_kaluki_ngilu) / sum(valid_votes) * 100, 1) 
  ) %>% 
  pivot_longer( 
    cols = c(Malombe, Mueke, Musila, Ngilu), 
    names_to  = 'candidate', 
    values_to = 'share' 
  ) 
 
# Build heatmap 
ggplot(share_matrix, aes(x = candidate, y = constituency_name, fill = share)) + 
  geom_tile(color = 'white', linewidth = 0.5) + 
  geom_text(aes(label = paste0(share, '%')), 
            size = 3.5, fontface = 'bold', color = 'white') + 
  scale_fill_gradient2( 
    low = '#EFF6FF', mid = '#3B82F6', high = '#1B3A6B', 
    midpoint = 25 
  ) + 
  labs( 
    title    = 'Vote Share Heatmap — 2022 Kitui Governor', 
    subtitle = 'Color intensity = higher vote share', 
    x = 'Candidate', y = 'Constituency', 
    fill = 'Vote Share (%)' 
  ) + 
  theme_minimal(base_size = 12) + 
  theme(plot.title = element_text(face = 'bold', color = '#1B3A6B')) 
 
# Explanation: 
# geom_tile()         draws colored rectangles for each cell 
# scale_fill_gradient2  two-tone gradient; mid sets the midpoint color 
# geom_text on tile   overlays the percentage value on each cell 
```








#Step 4.3 — Stacked Bar Chart: Vote Composition per Constituency

```{r}
 
ggplot(constituency_long, 
aes(x = constituency_name, y = votes, fill = candidate)) + 
geom_col(position = 'fill') +        # 'fill' makes each bar = 100% 
scale_y_continuous(labels = scales::percent) +  # Y axis as % 
scale_fill_manual(values = c( 
'Malombe' = '#1B3A6B', 'Mueke' = '#0EA5E9', 
'Musila'  = '#16A34A', 'Ngilu' = '#D97706' 
)) + 
coord_flip() +                       
# Flip to horizontal bars 
labs( 
title    = 'Vote Share Composition by Constituency', 
subtitle = 'Each bar sums to 100% of valid votes in that constituency', 
x = NULL, y = 'Vote Share' 
) + 
theme_minimal(base_size = 12) + 
theme( 
legend.position  = 'bottom', 
legend.title     = element_blank() 
) 
# Explanation: 
# position = 'fill'  stacks bars and normalizes each to 100% 
# coord_flip()       rotates the chart 90 degrees for readability 
# legend.title = element_blank()  removes the legend title 

```










#WEEK 5 — Polling Station Micro-Intelligence 
This week drills down to the individual polling station level to identify Mueke's best and worst performance 
locations — the building blocks of grassroots campaign targeting.
##Step 5.1 — Top & Bottom Performing Stations for Mueke
```{r}
# Top 15 polling stations by Mueke's vote share 
top_mueke <- df_clean %>% 
  filter(valid_votes >= 50) %>%
  select(constituency_name,polling_station_code,polling_station_name,votes_cast,jonathan_mwangangi_mueke,mueke_share)%>%
  arrange(desc(mueke_share)) %>% 
  head(15) 
 
# Bottom 15 (where Mueke got almost no votes) 
bottom_mueke <- df_clean %>% 
  filter(valid_votes >= 50) %>%
  select(constituency_name,polling_station_code,polling_station_name,votes_cast,jonathan_mwangangi_mueke,mueke_share)%>%
  arrange(mueke_share) %>% 
  head(15) 

View(top_mueke)
View(bottom_mueke)

# Explanation: 
# filter(valid_votes >= 50)  removes tiny stations where % is unreliable 
# arrange(desc(...))         sorts by highest share first 
# arrange(...)               without desc() = ascending (lowest first) 
# head(15)                   keeps only first 15 rows 
```









#Step 5.2 — Lollipop Chart: Mueke's Best Stations 
A lollipop chart is more elegant than a bar chart for ranked lists: 
```{r}

ggplot(bottom_mueke, 
       aes(x = reorder(polling_station_name, mueke_share), 
           y = mueke_share)) + 
  geom_segment(                            # Draw the 'stick' 
    aes(xend = polling_station_name, y = 0, yend = mueke_share), 
    color = '#0EA5E9', linewidth = 1 
  ) + 
  geom_point(                              # Draw the 'lollipop head' 
    size = 4, color = '#1B3A6B', fill = '#0EA5E9', 
    shape = 21, stroke = 1.5 
  ) + 
  geom_text(aes(label = paste0(round(mueke_share, 1), '%')), 
            hjust = -0.3, size = 3) + 
  coord_flip() + 
  labs( 
    title    = "Mueke's Bottoms 15 Polling Stations — 2022", 
    subtitle = 'Stations with at least 50 valid votes cast', 
    x = NULL, y = 'Mueke Vote Share (%)' 
  ) + 
  theme_minimal(base_size = 11) 
 
# Explanation: 
# geom_segment()   draws a line from (y=0) to (yend=mueke_share) 
# geom_point()     draws a circle at the end of each segment 
# shape = 21       a filled circle with a separate fill and stroke color 
# hjust = -0.3     moves text label slightly right of the point 
# reorder()        sorts the y-axis by mueke_share value 
```






#Step 5.3 — Scatter Plot: Vote Size vs Mueke Performance 
A scatter plot reveals whether Mueke performed better at large or small polling stations: 

```{r}

 
ggplot(df_clean, aes(x = valid_votes, y = mueke_share)) + 
  geom_point( 
    aes(color = constituency_name, size = valid_votes), 
    alpha = 0.5                           # Semi-transparent points 
  ) + 
  geom_smooth(method = 'lm', se = TRUE,  # Add linear trend line 
              color = '#1B3A6B', linewidth = 1) + 
  scale_color_viridis_d(option = 'D') +  # Color-blind-safe palette 
  scale_size(range = c(1, 5)) + 
  labs( 
    title    = 'Station Size vs Mueke Vote Share', 
    subtitle = 'Each point = one polling station; line = linear trend', 
    x        = 'Valid Votes Cast at Station', 
    y        = 'Mueke Vote Share (%)', 
    color    = 'Constituency' 
  ) + 
  theme_minimal(base_size = 12) + 
  guides(size = 'none') 
 
# Explanation: 
# geom_point()   draws one point per polling station 
# alpha = 0.5    50% transparency avoids overplotting 
# geom_smooth(method='lm')  fits and draws a linear regression line 
# se = TRUE      shows the 95% confidence interval band 
# scale_color_viridis_d   discrete viridis color palette 
# guides(size='none')     hides the size legend (size = valid_votes) 
```

































#WEEK 6 — Competitor Intelligence & Coalition Analysis 
This week students compare all four candidates and identify where voter transfers could boost Mueke in 2027 
— the foundation of coalition strategy. 
 
##Step 6.1 — Candidate Comparison Across Constituencies 
```{r} 
# Reshape to compare all candidates side by side 
comparison <- df_clean %>% 
  group_by(constituency_name) %>% 
  summarise(across( 
    c(julius_makau_malombe, jonathan_mwangangi_mueke, 
      david_musila, charity_kaluki_ngilu), 
    sum 
  )) %>% 
  rename( 
    Malombe = julius_makau_malombe, 
    Mueke   = jonathan_mwangangi_mueke, 
    Musila  = david_musila, 
    Ngilu   = charity_kaluki_ngilu 
  ) %>% 
  pivot_longer(-constituency_name, names_to='candidate', values_to='votes') 
 
# Explanation: 
# across()  applies the same function (sum) to multiple columns at once 
# rename()  renames columns for cleaner output 
# pivot_longer(-constituency_name)  pivots all columns except that one 
 
```


#Step 6.2 — Grouped Bar Chart: Head-to-Head Comparison 
```{r}

 
ggplot(comparison, 
       aes(x = constituency_name, y = votes, fill = candidate)) + 
  geom_col(position = 'dodge', width = 0.75) + # Bars side by side 
  scale_fill_manual(values = c( 
    'Malombe' = '#1B3A6B', 'Mueke' = '#0EA5E9', 
    'Musila'  = '#16A34A', 'Ngilu' = '#D97706' 
  )) + 
  scale_y_continuous(labels = scales::comma) + 
  labs( 
    title    = 'Candidate Votes by Constituency — 2022 Kitui Governor', 
    subtitle = 'Grouped bars show relative performance across all six 
constituencies', 
    x = NULL, y = 'Total Votes', fill = 'Candidate' 
  ) + 
  theme_minimal(base_size = 11) + 
  theme( 
    axis.text.x    = element_text(angle = 25, hjust = 1), 
    legend.position = 'top' 
  ) 
 
# Explanation: 
# position = 'dodge'  places bars for each candidate side-by-side 
#                     (vs 'stack' which stacks them) 
# legend.position='top'  moves legend above the chart 
 
```

#Step 6.3 — Correlation Heatmap: Candidate Vote Relationships 
A correlation matrix reveals whether candidates are drawing from the same voter pool — critical for coalition 
analysis: 

```{r} 
# Build correlation matrix from polling station votes 
vote_matrix <- df_clean %>% 
  select(julius_makau_malombe, jonathan_mwangangi_mueke, 
         david_musila, charity_kaluki_ngilu) 
 
cor_matrix <- cor(vote_matrix, use = 'complete.obs') 
 
# Rename for display 
colnames(cor_matrix) <- rownames(cor_matrix) <- 
  c('Malombe', 'Mueke', 'Musila', 'Ngilu') 
 
# Visualize with corrplot 
corrplot( 
  cor_matrix, 
  method  = 'color',           # Colored squares 
  type    = 'upper',           # Show upper triangle only 
  addCoef.col = 'black',       # Add correlation values 
  tl.col  = '#1B3A6B',         # Text label color 
  col     = colorRampPalette(c('#EFF6FF', '#3B82F6', '#1B3A6B'))(200), 
  title   = 'Candidate Vote Correlation — Polling Station Level', 
  mar     = c(0,0,2,0) 
) 
 
# Explanation: 
# cor()           calculates Pearson correlation between columns 
# use='complete.obs'  ignores rows with any NA values 
# method='color'  fills each cell with a color gradient 
# type='upper'    only shows the top-right triangle (avoids duplication) 
# addCoef.col     overlays the actual correlation coefficient value 
# 
# Correlation values range from -1 to +1: 
# +1.0 = perfect positive correlation 
#  0.0 = no correlation 
# -1.0 = perfect negative correlation

```
























#WEEK 7 — Voter Bloc Analysis & Distribution Plots 
This week students use distribution visualizations to understand the spread of Mueke's support across polling 
stations — identifying under-engaged zones and persuadable blocs. 
 
##Step 7.1 — Distribution of Mueke's Vote Share 
A histogram shows how frequently each vote share range occurs: 
 
```{r}
# Histogram of Mueke vote share across all polling stations 
ggplot(df_clean, aes(x = mueke_share)) + 
  geom_histogram( 
    binwidth = 1,               # Each bin covers 1 percentage point 
    fill = '#0EA5E9', 
    color = 'white', 
    alpha = 0.85 
  ) + 
  geom_vline( 
    xintercept = mean(df_clean$mueke_share, na.rm = TRUE), 
    linetype = 'dashed', color = '#1B3A6B', linewidth = 1 
  ) + 
  annotate('text', x = mean(df_clean$mueke_share) + 0.8, 
           y = 80, label = paste0('Mean: ', 
           round(mean(df_clean$mueke_share), 1), '%'), 
           color = '#1B3A6B', fontface = 'bold') + 
  labs( 
    title    = 'Distribution of Mueke Vote Share — Polling Station Level', 
    subtitle = 'Each bar = number of polling stations in that vote-share range', 
    x = 'Mueke Vote Share (%)', y = 'Number of Polling Stations' 
  ) + 
  theme_minimal(base_size = 12) 
 
# Explanation: 
# geom_histogram()   divides x axis into bins, counts rows in each bin 
# binwidth = 1       each bin covers 1 percentage point 
# geom_vline()       draws a vertical line at the mean 
# xintercept         the x-axis value where the line crosses 
# annotate('text')   adds a custom text label at specific coordinates 
```
 
##Step 7.2 — Box Plots: Vote Share Distribution by Constituency 
Box plots show the median, quartiles, and outliers — much more informative than just averages: 
 
```{r}
ggplot(df_clean, 
       aes(x = reorder(constituency_name, mueke_share, FUN = median), 
           y = mueke_share, fill = constituency_name)) + 
  geom_boxplot(alpha = 0.7, outlier.shape = 21, 
               outlier.fill = '#DC2626', outlier.size = 2) + 
  geom_jitter(width = 0.2, alpha = 0.2,  # Show all data points 
              color = '#1B3A6B', size = 0.8) + 
  scale_fill_viridis_d(option = 'C') + 
  coord_flip() + 
  labs( 
    title    = "Mueke Vote Share Distribution by Constituency", 
    subtitle = "Box = 25th–75th percentile; line = median; dots = individual 
stations", 
    x = NULL, y = 'Mueke Vote Share (%)' 
  ) + 
  theme_minimal(base_size = 12) + 
  theme(legend.position = 'none') 
 
# Explanation: 
# geom_boxplot()    shows: box (IQR), median line, whiskers, outliers 
# outlier.shape=21  outliers shown as filled circles 
# geom_jitter()     shows individual points with small random horizontal offset 
#                   to prevent overlap — reveals data density 
# reorder(..., FUN=median)  sorts constituencies by their median mueke_share 
# scale_fill_viridis_d  sequential color scale (option C = plasma) 
```




```{r}
library(forcats)
# Step 1: Calculate median mueke_share per constituency
medians <- tapply(df_clean$mueke_share, df_clean$constituency_name, median, na.rm = TRUE)

# Step 2: Reorder constituency_name factor by those medians
df_clean$constituency_name <- factor(
  df_clean$constituency_name,
  levels = names(sort(medians))
)

# Step 3: Now plot — no reorder() needed inside ggplot at all
ggplot(df_clean,
       aes(x = constituency_name,
           y = mueke_share,
           fill = constituency_name)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21,
               outlier.fill = '#DC2626', outlier.size = 2) +
  geom_jitter(width = 0.2, alpha = 0.2,
              color = '#1B3A6B', size = 0.8) +
  scale_fill_viridis_d(option = 'C') +
  coord_flip() +
  labs(
    title    = "Mueke Vote Share Distribution by Constituency",
    subtitle = "Box = 25th-75th percentile; line = median; dots = individual stations",
    x = NULL, y = 'Mueke Vote Share (%)'
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = 'none')
```
 
##Step 7.3 — Violin Plot: Comparing Vote Distribution Shapes 
```{r} 
# Reshape all four candidates for comparison 
all_shares <- df_clean %>% 
  select(constituency_name, malombe_share, mueke_share, 
         musila_share, ngilu_share) %>% 
  pivot_longer( 
    cols = ends_with('_share'), 
    names_to  = 'candidate', 
    values_to = 'share' 
  ) %>% 
  mutate(candidate = str_replace(candidate, '_share', '') %>% 
                     str_to_title()) 
 
ggplot(all_shares, aes(x = candidate, y = share, fill = candidate)) + 
  geom_violin(trim = FALSE, alpha = 0.7) + 
  geom_boxplot(width = 0.1, fill = 'white', alpha = 0.8) + 
  scale_fill_manual(values = c( 
    'Malombe_share' = '#1B3A6B', 'Mueke_share' = '#0EA5E9', 
    'Musila_share'  = '#16A34A', 'Ngilu_share' = '#D97706' 
  )) + 
  labs( 
    title    = 'Vote Share Distribution — All Candidates', 
    subtitle = 'Width of violin = density of polling stations at that share 
level', 
    x = NULL, y = 'Vote Share (%)' 
  ) + 
  theme_minimal(base_size = 12) + 
  theme(legend.position = 'none') 
 
# Explanation: 
# geom_violin()   shows the full distribution shape — wider = more stations 
#                 at that vote share level 
# trim = FALSE    shows the full tails of the distribution 
# Overlaid boxplot (width=0.1) adds summary stats inside the violin
```























#WEEK 8 — Rejected Votes & Turnout Intelligence 
Rejected votes and turnout patterns reveal campaign mobilization effectiveness and potential ballot integrity 
issues. High rejection rates often signal voter confusion or logistical failure. 
 
##Step 8.1 — Rejection Rate Analysis 

```{r} 
# Summary of rejection rates by constituency 
rejection_summary <- df_clean %>% 
  group_by(constituency_name) %>% 
  summarise( 
    total_rejected   = sum(rejected_votes), 
    total_cast       = sum(votes_cast), 
    rejection_rate   = round((total_rejected / total_cast) * 100, 2), 
    avg_station_rate = round(mean(rejection_rate), 2) 
  ) %>% 
  arrange(desc(rejection_rate)) 
 
print(rejection_summary) 
 
# Bar chart of rejection rates 
ggplot(rejection_summary, 
       aes(x = reorder(constituency_name, rejection_rate), 
           y = rejection_rate, fill = rejection_rate)) + 
  geom_col(width = 0.6) + 
  geom_text(aes(label = paste0(rejection_rate, '%')), 
            hjust = -0.2, size = 4, fontface = 'bold') + 
  scale_fill_gradient(low = '#BBF7D0', high = '#DC2626') + 
  coord_flip() + 
  labs( 
    title    = 'Rejected Vote Rate by Constituency — 2022', 
    subtitle = 'Higher rejection rates may indicate voter education gaps', 
    x = NULL, y = 'Rejection Rate (%)' 
  ) + 
  theme_minimal(base_size = 12) + 
  theme(legend.position = 'none') 
 
# Explanation: 
# scale_fill_gradient()  continuous color scale from low to high value 
#                        green (low rejection) to red (high rejection) 
# This chart answers: which constituencies need more voter education? 
 
```


#Step 8.2 — Scatter: Station Size vs Rejection Rate 

```{r} 
ggplot(df_clean, aes(x = votes_cast, y = rejection_rate)) + 
  geom_point(aes(color = constituency_name), alpha = 0.4, size = 2) + 
geom_smooth(method = 'loess', se = TRUE, 
color = '#DC2626', fill = '#FECACA') + 
scale_color_viridis_d() + 
labs( 
title    = 'Polling Station Size vs Rejection Rate', 
subtitle = 'loess smooth shows trend — do larger stations reject more?', 
x = 'Votes Cast at Station', y = 'Rejection Rate (%)', 
color = 'Constituency' 
) + 
theme_minimal(base_size = 12) 
# Explanation: 
# method = 'loess'   
#fits a LOCALLY WEIGHTED smooth curve (not straight line) 
#                    better for detecting non-linear relationships 
# This pattern matters: if large stations reject more, it suggests 
# overcrowding and voter confusion — targetable with education campaigns
```


















#WEEK 9 — Advanced Aggregation & Statistical Analysis 
#This week introduces advanced summarization, including proportional analysis, margin calculations, and the use 
#of dplyr's across() and window functions. 
 
Step 9.1 — Mueke's Performance Gap vs Winner

```{r}
# For each station, calculate Mueke's gap vs the station winner 
df_gap <- df_clean %>% 
  mutate( 
    station_winner_votes = pmax( 
      julius_makau_malombe, david_musila, charity_kaluki_ngilu 
    ), 
    mueke_gap = jonathan_mwangangi_mueke - station_winner_votes, 
    gap_pct   = (mueke_gap / valid_votes) * 100 
  ) 
 
# Density plot of the gap 
ggplot(df_gap, aes(x = gap_pct, fill = constituency_name)) + 
  geom_density(alpha = 0.4) + 
  geom_vline(xintercept = 0, linetype = 'dashed', color = 'black') + 
  scale_fill_viridis_d() + 
  facet_wrap(~ constituency_name, ncol = 2) + 
  labs( 
    title    = "Mueke's Vote Gap vs Station Winner — by Constituency", 
    subtitle = 'Negative = Mueke behind; Positive = Mueke ahead at that 
station', 
    x = 'Mueke Gap vs Winner (% of valid votes)', y = 'Density' 
  ) + 
  theme_minimal(base_size = 11) + 
  theme(legend.position = 'none') 
 
# Explanation: 
# geom_density()   draws a smooth estimate of the distribution 
#                  (like a smoothed histogram) 
# pmax() without Mueke's own column = the best opponent score 
# Density to the RIGHT of 0 = stations where Mueke actually won
```



#Step 9.2 — Cumulative Vote Share Analysis

```{r}
# Sort stations by votes cast and calculate cumulative Mueke share 
df_cumulative <- df_clean %>% 
  arrange(votes_cast) %>% 
  mutate( 
    cumulative_mueke = cumsum(jonathan_mwangangi_mueke), 
    cumulative_valid = cumsum(valid_votes), 
    cumulative_share = (cumulative_mueke / cumulative_valid) * 100, 
station_rank     = row_number() 
) 
ggplot(df_cumulative, aes(x = station_rank, y = cumulative_share)) + 
geom_line(color = '#0EA5E9', linewidth = 1.2) + 
geom_hline(yintercept = mean(df_clean$mueke_share), 
linetype = 'dashed', color = '#DC2626') + 
labs( 
title    = 'Cumulative Mueke Vote Share as Stations Are Added', 
subtitle = 'Sorted by station size (smallest to largest)', 
x = 'Polling Station Rank (by size)', y = 'Cumulative Vote Share (%)' 
) + 
theme_minimal(base_size = 12) 
# Explanation: 
# cumsum()       running total (cumulative sum) 
# row_number()   assigns 1,2,3... to each row (useful as x-axis) 
# geom_line()    connects points with a line — good for trends over sequence 

```




















#WEEK 10 — Machine Learning: K-Means Clustering 
Clustering groups polling stations with similar political profiles together without needing labeled outcomes. This 
identifies natural political blocs — strongholds, swing zones, and lost territory. 
 
#$Step 10.1 — Prepare Features for Clustering 
```{r}
# Select features for clustering 
cluster_features <- df_clean %>% 
  select( 
    malombe_share, mueke_share, musila_share, ngilu_share, 
    rejection_rate, valid_votes 
  ) %>% 
  na.omit()                    # Remove any remaining NAs 
 
# Scale the features — REQUIRED for k-means 
cluster_scaled <- scale(cluster_features) 
 
# Explanation: 
# scale()   standardizes each column to mean=0, sd=1 
#           This is ESSENTIAL because variables on different scales 
#           (e.g., votes 0-500 vs percentages 0-100) would cause 
#           the larger-scale variable to dominate clustering 
# na.omit() removes rows with any missing values 
``` 
 
##Step 10.2 — Find Optimal Number of Clusters (Elbow Method) 

```{r} 
# Calculate within-cluster sum of squares for k = 1 to 10 
set.seed(42)   # Set seed for reproducibility 
wss <- sapply(1:10, function(k) { 
  kmeans(cluster_scaled, centers = k, nstart = 25)$tot.withinss 
}) 
 
# Plot the elbow curve 
elbow_df <- data.frame(k = 1:10, wss = wss) 
 
ggplot(elbow_df, aes(x = k, y = wss)) + 
  geom_line(color = '#1B3A6B', linewidth = 1.2) + 
  geom_point(size = 4, color = '#0EA5E9', fill = '#0EA5E9', shape = 21) + 
  geom_vline(xintercept = 4, linetype = 'dashed', color = '#DC2626') + 
  scale_x_continuous(breaks = 1:10) + 
  labs( 
    title    = 'Elbow Method — Optimal Number of Clusters', 
    subtitle = 'Look for the elbow: the k where WSS stops dropping sharply', 
    x = 'Number of Clusters (k)', y = 'Total Within-Cluster SS' 
  ) + 
  theme_minimal(base_size = 12) 
 
# Explanation: 
# set.seed(42)    makes random results reproducible 
# sapply()        applies a function over a vector, returns a vector 
# kmeans()        fits k-means clustering 
# centers = k     number of clusters to create 
# nstart = 25     tries 25 random starting points, keeps best result 
# $tot.withinss   total within-cluster sum of squares 
#                 (lower = more compact clusters) 
# The 'elbow' = the k value where improvement slows — best balance 
 
```


##Step 10.3 — Fit the K-Means Model 

```{r} 
# Fit k-means with k = 4 clusters 
set.seed(42) 
km_model <- kmeans(cluster_scaled, centers = 4, nstart = 25) 
 
# Add cluster assignment back to the original data 
df_clustered <- df_clean %>% 
  na.omit() %>% 
  mutate(cluster = as.factor(km_model$cluster)) 
 
# View cluster sizes 
table(df_clustered$cluster) 
 
# View cluster centers (interpret the profile of each cluster) 
km_model$centers 
 
# Explanation: 
# km_model$cluster  the cluster label (1-4) assigned to each row 
# as.factor()       converts numeric cluster IDs to categorical labels 
#                   important so ggplot treats them as groups, not numbers 
# $centers          the mean feature value for each cluster center 
#                   use this to interpret what each cluster represents 
 

# table() output example: 
# 1   2   3   4 
# 312 198 287 193 
 
# $centers interpretation (approximate): 
# Cluster 1: High Malombe, Low Mueke  -> Malombe strongholds 
# Cluster 2: High Musila, Low Malombe -> Musila territory 
# Cluster 3: Balanced split           -> Competitive / swing stations 
# Cluster 4: Low all candidates       -> Very small / low-turnout stations 
 
```


#Step 10.4 — Visualize Clusters with PCA 
Principal Component Analysis (PCA) reduces multi-dimensional data to 2D for plotting: 
```{r} 
# Run PCA manually
pca_result <- prcomp(cluster_scaled)

# Extract first 2 components
pca_df <- data.frame(
  PC1     = pca_result$x[, 1],
  PC2     = pca_result$x[, 2],
  cluster = as.factor(km_model$cluster)
)

# Plot manually with ggplot2
ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.6, size = 2) +
  stat_ellipse(linewidth = 0.8) +
  scale_color_manual(values = c(
    '1' = '#1B3A6B',
    '2' = '#0EA5E9',
    '3' = '#16A34A',
    '4' = '#D97706'
  )) +
  labs(
    title    = "K-Means Clusters of Kitui Polling Stations",
    subtitle = "PCA reduces 6 features to 2 dimensions for visualization",
    x        = "Principal Component 1",
    y        = "Principal Component 2",
    color    = "Cluster"
  ) +
  theme_minimal(base_size = 12) 
 
```

##Step 10.5 — Map Clusters Back to Constituencies 

```{r} 
# See which constituencies have most of each cluster type 
cluster_by_constituency <- df_clustered %>% 
  count(constituency_name, cluster) %>% 
  group_by(constituency_name) %>% 
  mutate(pct = n / sum(n) * 100) 
 
ggplot(cluster_by_constituency, 
       aes(x = constituency_name, y = pct, fill = cluster)) + 
  geom_col(position = 'fill') + 
  scale_y_continuous(labels = scales::percent) + 
  scale_fill_manual(values = c( 
    '1' = '#1B3A6B', '2' = '#0EA5E9', '3' = '#16A34A', '4' = '#D97706' 
  )) + 
  coord_flip() + 
  labs( 
    title    = 'Cluster Composition by Constituency', 
    subtitle = 'What proportion of each constituency belongs to each voter 
cluster?', 
    x = NULL, y = '% of Stations', fill = 'Cluster' 
  ) + 
  theme_minimal(base_size = 12)
```


















#WEEK 11 — Machine Learning: Predictive Model 
This week students build a Random Forest classifier to predict whether Mueke will win a polling station in 2027 
given station characteristics — the electability prediction model. 
 
##Step 11.1 — Create the Target Variable 

```{r} 
# Define 'mueke_competitive' as TRUE if Mueke gets > 10% share 
# This represents a meaningful threshold, not just winning 
df_model <- df_clean %>% 
  mutate( 
    mueke_competitive = as.factor(ifelse(mueke_share > 10, 'Yes', 'No')) 
  ) %>% 
  select( 
    mueke_competitive, 
    valid_votes, rejection_rate, 
    malombe_share, musila_share, ngilu_share, 
    constituency_name 
  ) %>% 
  na.omit() 
 
# Check class balance 
table(df_model$mueke_competitive) 
 
# Explanation: 
# ifelse(condition, yes_value, no_value)  vectorized if-else 
# as.factor()   converts Yes/No to a categorical factor 
#               caret requires the target to be a factor for classification 
# select()      keeps only the columns we want as model features 

```


 
##Step 11.2 — Split into Training and Test Sets 

```{r} 
library(caret)

set.seed(42) 
 
# Create 75/25 train/test split 
train_index <- createDataPartition( 
  df_model$mueke_competitive, 
  p    = 0.75, 
  list = FALSE 
) 
 
df_train <- df_model[train_index, ] 
df_test  <- df_model[-train_index, ] 
 
cat('Training rows:', nrow(df_train), '\n') 
cat('Test rows:    ', nrow(df_test), '\n') 
 
# Explanation: 
# createDataPartition()   caret function for stratified splitting 
#                         'stratified' means it preserves the Yes/No ratio 
# p = 0.75                75% goes to training 
# list = FALSE            returns a vector of row indices, not a list 
# df_model[-train_index,] the NEGATIVE index = all rows NOT in train_index 
 
```


##Step 11.3 — Train the Random Forest Model 

```{r} 
# Set up cross-validation 
train_control <- trainControl( 
  method  = 'cv',    # k-fold cross-validation 
  number  = 5,       # 5 folds 
  verboseIter = FALSE 
) 
 
# Train the Random Forest 
set.seed(42) 
rf_model <- train( 
  mueke_competitive ~ .,  # Target ~ all remaining columns as predictors 
  data      = df_train, 
  method    = 'rf',       # Random Forest algorithm 
  trControl = train_control, 
  importance = TRUE       # Calculate feature importance 
) 
 
print(rf_model) 
 
# Explanation: 
# trainControl()     configures how the model is validated during training 
# method='cv'        cross-validation: data is split into 5 folds, 
#                    the model trains on 4 and tests on 1, repeated 5 times 
# train()            caret's unified training function 
# mueke_competitive ~ .   the formula interface: target ~ all other columns 
# method='rf'        uses the randomForest package under the hood 
# importance=TRUE    enables variable importance calculation 

``` 
 

##Step 11.4 — Evaluate Model Performance 

```{r} 
# Make predictions on the test set 
predictions <- predict(rf_model, newdata = df_test) 
 
# Confusion matrix 
conf_matrix <- confusionMatrix(predictions, df_test$mueke_competitive) 
print(conf_matrix) 
 
# Explanation of confusion matrix output: 
# Accuracy      = (TP + TN) / Total — overall correct predictions 
# Sensitivity   = TP / (TP + FN)    — how well we find 'Yes' cases 
# Specificity   = TN / (TN + FP)    — how well we find 'No' cases 
# Kappa         = agreement beyond chance (>0.6 is good) 
 
# Extract key metrics 
cat('Accuracy: ', round(conf_matrix$overall['Accuracy'] * 100, 1), '%\n') 
cat('Kappa:    ', round(conf_matrix$overall['Kappa'], 3), '\n') 
 
# Confusion Matrix output: 
#           Reference 
# Prediction  No  Yes 
#         No  178  12 
#         Yes   8  50 
 
# Accuracy: ~91.4% 
# Kappa: 0.782 (strong agreement) 

```


##Step 11.5 — Variable Importance Plot 

```{r} 
# Extract variable importance safely
importance_df <- varImp(rf_model)$importance

# View actual column names
print(names(importance_df))

# Convert row names to column
importance_df <- importance_df %>%
  rownames_to_column('variable')

# Rename the first importance column to 'importance'
# regardless of what it is called
names(importance_df)[2] <- 'importance'

# Arrange by importance
importance_df <- importance_df %>%
  arrange(desc(importance))

# Plot
ggplot(importance_df,
       aes(x = reorder(variable, importance),
           y = importance,
           fill = importance)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = round(importance, 1)),
            hjust = -0.2, size = 3.5) +
  scale_fill_gradient(low = '#BFDBFE', high = '#1B3A6B') +
  coord_flip() +
  labs(
    title    = 'Random Forest Variable Importance',
    subtitle = 'Higher score = stronger predictor of Mueke competitiveness',
    x = NULL, y = 'Importance Score'
  ) +
  theme_minimal(base_size = 12)

```

















#WEEK 12 — Final Dashboard Assembly & Scenario Modeling 
The final week brings all analyses together into a combined political intelligence dashboard and runs the three 
electability scenarios from the training plan. 

 
##Step 12.1 — Multi-Panel Dashboard with patchwork 
 
```{r}
# Install patchwork for combining ggplot2 charts 
#install.packages('patchwork') 
library(patchwork) 
 
# Create individual plots (abbreviated — use full versions from earlier weeks) 
p1 <- ggplot(county_totals, aes(x=reorder(candidate,-votes), y=votes, 
fill=candidate)) + 
        geom_col() + scale_fill_manual(values=c( 
          'Malombe'='#1B3A6B','Mueke'='#0EA5E9', 
          'Musila'='#16A34A','Ngilu'='#D97706')) + 
        labs(title='Total Votes', x=NULL, y='Votes') + 
        theme_minimal() + theme(legend.position='none') 
 
p2 <- ggplot(df_clean, aes(x=mueke_share)) + 
        geom_histogram(binwidth=1, fill='#0EA5E9', color='white') + 
        labs(title='Mueke Share Distribution', x='%', y='Stations') + 
        theme_minimal() 
 
p3 <- ggplot(constituency_summary, 
             aes(x=reorder(constituency_name, mueke_pct), y=mueke_pct)) + 
        geom_col(fill='#0EA5E9') + coord_flip() + 
        labs(title='Mueke % by Constituency', x=NULL, y='%') + 
        theme_minimal() 
 
p4 <- ggplot(df_clean, aes(x=valid_votes, y=mueke_share, 
color=constituency_name)) + 
        geom_point(alpha=0.3, size=1) + 
        geom_smooth(method='lm', se=FALSE, color='#1B3A6B') + 
        scale_color_viridis_d() + 
        labs(title='Size vs Mueke Share', x='Valid Votes', y='Mueke %') + 
        theme_minimal() + theme(legend.position='none') 
 
# Combine into 2x2 dashboard 
dashboard <- (p1 + p2) / (p3 + p4) + 
  plot_annotation( 
    title    = 'Jonathan Mueke 2022 Election Intelligence Dashboard', 
    subtitle = 'Source: IEBC Form 37C — Kitui County Gubernatorial Results', 
    theme    = theme(plot.title = element_text(size=16, face='bold', 
color='#1B3A6B')) 
  ) 
 
ggsave('mueke_dashboard.png', dashboard, width=14, height=10, dpi=300)
print(dashboard)
 
# Explanation: 
# patchwork uses + to place plots side by side 
# patchwork uses / to stack plots vertically 
# (p1+p2)/(p3+p4)  creates a 2x2 grid 
# plot_annotation() adds a single title/caption to the whole composite 
 

```



##Step 12.2 — Scenario Modeling (2027 Electability) 
The three scenarios from the training plan model how many votes Mueke needs to become competitive: 
 
```{r}
# Current baseline from 2022 results 
baseline <- sum(df_clean$jonathan_mwangangi_mueke) 
total_valid_2022 <- sum(df_clean$valid_votes) 
 
# Scenario parameters 
scenarios <- data.frame( 
  scenario      = c('A: Low Growth', 'B: Medium Growth', 'C: High Growth'), 
  growth_factor = c(1.5, 2.5, 4.0),    # Multiplier on 2022 votes 
  description   = c('Remains non-competitive', 
                     'Becomes coalition factor', 
                     'Viable gubernatorial contender') 
) 
 
scenarios <- scenarios %>% 
  mutate( 
    projected_votes = round(baseline * growth_factor), 
    projected_share = round((projected_votes / total_valid_2022) * 100, 1) 
  ) 
 
print(scenarios) 
 
# Visualize scenarios 
ggplot(scenarios, 
       aes(x = scenario, y = projected_share, fill = scenario)) + 
  geom_col(width = 0.6) + 
  geom_hline(yintercept = c(10, 20, 30), linetype='dashed', 
             color=c('#16A34A','#D97706','#DC2626'), linewidth=0.8) + 
  geom_text(aes(label = paste0(projected_share, '%')), 
            vjust = -0.5, fontface='bold', size=4.5) + 
  annotate('text', x=0.5, y=10.5, label='10%: Coalition Threshold', 
           hjust=0, size=3, color='#16A34A') + 
  annotate('text', x=0.5, y=20.5, label='20%: Serious Contender', 
           hjust=0, size=3, color='#D97706') + 
  annotate('text', x=0.5, y=30.5, label='30%: Competitive Race', 
           hjust=0, size=3, color='#DC2626') + 
  scale_fill_manual(values = c( 
    'A: Low Growth'    = '#94A3B8', 
    'B: Medium Growth' = '#0EA5E9', 
    'C: High Growth'   = '#1B3A6B' 
  )) + 
  scale_y_continuous(limits = c(0, 45)) + 
  labs( 
    title    = '2027 Mueke Electability Scenarios', 
    subtitle = 'Projected vote share under three growth assumptions', 
    x = 'Scenario', y = 'Projected Vote Share (%)' 
  ) + 
  theme_minimal(base_size = 12) + 
  theme(legend.position='none') 
 
ggsave('week12_scenarios.png', width=9, height=6, dpi=300) 
 
# Explanation: 
# data.frame()   creates a new data frame from scratch 
# Multiple geom_hline()  draws reference lines at key thresholds 
# annotate()     adds labels at exact x,y coordinates 
# This chart directly answers the training plan's Scenario A/B/C analysis 
 
# Scenario output (approximate): 
# A: Low Growth    — 1.5x -> ~27,000 votes -> ~6.0% — Non-competitive 
# B: Medium Growth — 2.5x -> ~45,000 votes -> ~10.0% — Coalition factor 
# C: High Growth   — 4.0x -> ~72,000 votes -> ~16.1% — Serious contender 
 
```





##Step 12.3 — Export Final Analysis Summary 

```{r} 
# Create a final summary for reporting 
final_summary <- df_clean %>% 
  group_by(constituency_name) %>% 
  summarise( 
    stations         = n(), 
    mueke_total      = sum(jonathan_mwangangi_mueke), 
    mueke_pct        = round(mean(mueke_share, na.rm=TRUE), 2), 
    stations_won     = sum(winner == 'Mueke'), 
    pct_won          = round(stations_won / stations * 100, 1), 
    avg_rejection    = round(mean(rejection_rate, na.rm=TRUE), 2) 
  ) %>% 
  arrange(desc(mueke_pct)) 
 
# Export to CSV 
write.csv(final_summary, 'mueke_final_intelligence.csv', row.names=FALSE) 
 
# Export full cleaned dataset 
write.csv(df_clean, 'kitui_full_clean.csv', row.names=FALSE) 
 
# Export the clustered dataset 
write.csv(df_clustered, 'kitui_clustered.csv', row.names=FALSE) 
 
# Print final summary 
cat('=== JONATHAN MUEKE 2022 BASELINE INTELLIGENCE ===\n') 
cat('Total votes received:', sum(df_clean$jonathan_mwangangi_mueke), '\n') 
cat('County vote share:   ', round(mean(df_clean$mueke_share), 2), '%\n') 
cat('Polling stations won:', sum(df_clean$winner=='Mueke'), 'of', 
nrow(df_clean), '\n') 
cat('Best constituency:  ', final_summary$constituency_name[1], '\n') 
cat('Weakest area:       ', tail(final_summary$constituency_name, 1), '\n') 
 

# === JONATHAN MUEKE 2022 BASELINE INTELLIGENCE === 
# Total votes received: ~18,000 
# County vote share:    
#~4.0% 
# Polling stations won: ~42 of 990 
# Best constituency:    
#MWINGI NORTH 
# Weakest area:         
#KITUI WEST 

```








